Poker Chips, Earthquakes, and Computation

### 2014-10-11 finaldraft.tex
modified half of stability section, up to bound on board size

The motivation for this definition is that the spread function should give a measure of how ``spread out" the chips on a board are. Intuitvely, earthquakes have the function of moving chips that are condensed in a single piile and spreading them out over a greater area. This function, which sums up the chip stack sizes weighted by the taxicab distance squared to the origin, sufficiently captures the idea of spread.

\begin{theorem}
\label{finitestability}
If a board $b$ is finite then there exists $N>0$ such that after applying a $N$ earthquakes the resultant board will be stable.
\end{theorem}

In order to show finite boards never revisit a prior arrangement, we define a spread function that maps boards to nonnegative integers. The spread function should give some measure of how ``spread out" the chips on a board are. Earthquakes seem to ``spread out" the chips on a board and so we would like earthquakes to increase the value of the function.

\begin{lemma}
Suppose $b$ is a finite board. If $E_{s}b \neq b$, then no application of earthquakes to $E_{s}b$ can produce $b$. 
\end{lemma}

\begin{proof}
For ease of notation we use $s=(s_1, s_2) \in \mathbb{Z} \times \mathbb{Z}$. We define our spread function as a map from finite boards to the nonnegative integers
\begin{equation*}
S(b) = \sum_{i,j} b(i,j)(|i|+|j|)^2. 
\end{equation*}
We can think of $S(b)$ as the sum of the number of chips on each square weighted by the "taxi-cab" distance from the origin squared. This is well-defined since the total number of chips on the board is finite. We claim that 
\begin{equation*}
E_{s} b \neq b \Rightarrow S(E[b, s]) > S(b)
\end{equation*}
That is if an earthquake changes the board, $S$ increases. This implies that if an earthquake changes a board, then no subsequent application of earthquakes can transform it back to its original state.

We evaluate $S(E_{s} b) - S(b)$, recalling that $E_s b(i,j)$ refers to the value of $b$ at $(i,j)$ after an earthquake effected $s$.
\begin{align*}
S(E_{s} b) - S(b) &=\sum_{i,j} E_{s} b (i,j)(|i|+|j|)^2 - \sum_{i,j} b(i,j)(|i|+|j|)^2 \\
&=-4(|s_1| + |s_2|)^2 +\sum_{(i,j) \in n(s)} (|i| + |j|)^2
\end{align*}
For at most two of the neighbors of $(s_1, s_2)$ 
\begin{equation*}
(|i|+|j|)= (|s_1|+|s_2|-1)^2,
\end{equation*}
and for at least two of the neighbors 
\begin{equation*}
(|i|+|j|)=(|s_1|+|s_2|+1)^2,
\end{equation*}
 and $(|s_1|+|s_2|+1)^2 > (|s_1|+|s_2|-1)^2$. Therefore 
\begin{align*}
&-4(|s_1| + |s_2|)^2 +\sum_{(i,j) \in n(s)} (|i| + |j|)^2 \\
& \geq 4(|s_1| + |s_2|)^2 + 2 (|s_1| + |s_2|+1)^2 + 2 (|s_1| + |s_2|-1)^2 = 4 >0.
\end{align*}
\end{proof}

>>>>

Thus, the function sums up the chip stack sizes weighted by the taxicab distance to the origin squared. The motivation for this definition is that the spread function gives a measure of how ``spread out" the chips on a board are. Intuitvely, earthquakes move chips that are condensed in a single pile and spreads them out over a greater area. We capture this in the following lemma:

\begin{lemma}
\label{spreadfunction}
Suppose an earthquake hits node $s$ of a board $b$, producing the board $E_sb$. Then
\[ \Psi(E_sb) > \Psi(b) \]
so the earthquake strictly increases the spread of the board.
\end{lemma}
\begin{proof}
The chip stacks of the two boards are the same at all nodes except at $s$ and $n(s)$, the nodes that neighbor $s$. Thus, the contribution to $\Psi$ is the same at all nodes except for those 5.

Let the taxicab distance to $s$ is $d_s$. Then the contribution to the spread function from node $s$ decreases from $b(s)d_s^2$ to $(b(s)-4)d_s^2$.

Out of the 4 nodes in $n(s)$, some have a taxicab distance of $d_s+1$, and the rest are at distance $d_s-1$. In particular, if $s$ is not exactly on the $x$ or $y$ axis, exactly 2 neighboring nodes are at $d_s+1$, and exactly 2 are at distance $d_s-1$. If $s$ is on either the $x$ or $y$ axis, then 3 nodes are at distance $d_s+1$, and 1 neighboring node is at distance $d_s-1$. Since we want to lower bound the increase in the spread function, we focus only on the case where exactly 2 neighboring nodes are at distance $d_s+1$, as it can be shown that this case leads to a strictly smaller increase than the other.

Suppose that the two neighboring nodes at distance $d_s+1$ have chip stacks of $n_1$ and $n_2$ originally. Then their contribution to the spread function goes from $n_1(d_s+1)^2$ and $n_2(d_s+1)^2$ to $(n_1+1)(d_s+1)^2$ and $(n_2+1)(d_s+1)^2$. If the other two nodes at distance $d_s-1$ have chip stacks of $n_3$ and $n_4$, their contribution to the spread function goes from $n_3(d_s-1)^2$ and $n_4(d_s-1)^2$ to $(n_3+1)(d_s-1)^2$ and $(n_4+1)(d_s-1)^2$.

Omitting the algebraic details, we find that summing up the changes in the spread function gives
\[ \Psi(E_sb) - \Psi(b) = 4\]
which completes the proof.
\end{proof}

\begin{corollary}
\label{finitestability}
A board can never return to a previous state via a sequence of earthquakes.
\end{corollary}
\begin{proof}
Suppose for the sake of contradiction that this is possible, and that $b = E_1E_2\cdots E_kb$ for some sequence of earthquakes $E_1,E_2,\dots,E_k$. Lemma~\ref{spreadfunction} shows that $\psi(b) > \psi(E_1E_2\cdots E_kb)$, which is a contradiction.
\end{proof}
